Rényi Entropy-Based Heterogeneity Detection in SAR Data

### 2025-03-15 Text/Articles/RENYI_ENTROPY_SAR_2025/R0-Renyi-entropy-sar.qmd
Update paper with corrected comments

---
title: Rényi Entropy-Based Heterogeneity Detection in SAR Data
format:
  ieee-pdf:
    pdf-engine:  pdflatex # xelatex
    keep-tex: true  
    conference: true # comment this line to use journal
    journaltype: conference # comment this line to use journal
    fig-cap-location: bottom # to crossref figure
   # link-citations: true
    #colorlinks: true
    #linkcolor: blue
  ieee-html: default

author:
  - id: 
    name: Janeth Alpala, Abraão D.\ C.\ Nascimento 
    affiliations:
      - name: Universidade Federal de Pernambuco
        department: Departamento de Estatística
        city: Recife 
        country: PE, Brazil
        postal-code: 18800
      - name: Unknown affiliation
    #orcid: 
    email: janeth.alpala@ufpe.br, abraao@de.ufpe.br
    #url: 
    #membership: "Member, IEEE"
    attributes:
      corresponding: true
    #photo: 
    #bio: |
    #  Use `IEEEbiography`  with figure as  option and
    #  the author name as the argument followed by the biography text.
  - name: Alejandro C.\ Frery
    affiliations:
      - name: Victoria University of Wellington
        department: School of Mathematics and Statistics
        city: Wellington 
        country: New Zealand
        postal-code: 18800
      - name: Unknown affiliation
    #orcid: 
    email: alejandro.frery@vuw.ac.nz
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems. A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes. In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed. 
  In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task. 
  In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy. 
  In particular, we use a nonparametric estimator for the Rényi entropy and investigate some of its properties. 
  As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories. 
  The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy.
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR,  hypothesis tests]
#funding: 
funding: 
  statement: "The `quarto-ieee` template is freely available under the MIT license on github: <https://github.com/dfolio/quarto-ieee>."
pageheader:
  left: Journal XXX, Month Year
  right: #'D. Folio:  A Sample Article Using quarto-ieee'

header-includes:
   - \usepackage[english]{babel}
   - \usepackage{bm,bbm}
   - \usepackage{mathrsfs}
   - \usepackage{siunitx}
   - \usepackage{graphicx}
   - \usepackage{url}
   - \usepackage[T1]{fontenc}
   - \usepackage{polski}
   - \usepackage{booktabs}
   - \usepackage{color}
   - \usepackage{mathtools}
   - \usepackage[utf8]{inputenc}
   - \usepackage{hyperref}
   - \hypersetup{draft}
   - \usepackage{float}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{xcolor}
   - \usepackage{amsmath}

#bibliography: references.bib
bibliography: ../../Common/references.bib
#

@@

---

```{r setup, include=FALSE}

@@

<!-- \newcommand{\reallywidehat}[1]{\widehat{#1}} -->

# Introduction

Synthetic Aperture Radar (SAR) technology has become essential for a wide range of applications\ [@Yu2023;@Akbarizadeh2012;@Mondini2021]. It provides high-resolution data that is independent of sunlight and operates in a variety of weather conditions to facilitate global Earth monitoring\ [@Zeng2020].
Despite its advantages, the effective use of SAR data depends on understanding its statistical properties because this data is affected by speckle, a noise-like interference effect\ [@Argenti2013; @Choi2019]. In intensity format, speckle is typically non-Gaussian, and its presence complicates subsequent image analysis tasks.

The $\mathcal{G}^0$ distribution has been shown to be an effective model for SAR intensity data, as it captures different levels of roughness. A special case is the Gamma distribution, which arises when speckle is fully developed, indicating homogeneous regions. 
Although these models offer flexibility in describing SAR intensities, selecting the appropriate model can be challenging for two main reasons: the small sample sizes used in practical applications and the inherent difficulties associated with parameter estimation. 
These challenges complicate model selection and highlight the need to explore alternative statistical approaches.

Entropy measures have gained attention as valuable statistical tools for analyzing SAR data, with applications in edge detection\ [@Nascimento2014], segmentation\ [@Nobre2016], classification\ [@Cassetti2022], and noise reduction\ [@Chan2022].
Traditionally, Shannon entropy\ [@Shannon1948] has been widely used to quantify uncertainty and disorder in data. However, this study explores an alternative information measure: Rényi entropy. 
As a generalization of Shannon entropy, it offers additional insights for identifying heterogeneity, making it a promising tool for enhancing SAR image analysis.

@@

The main models considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which is able to describe roughness\ [@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{align}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    &= \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
    \intertext{and}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    &= \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}
    \notag \\
    &\quad \times
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{align}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
As demonstrated by\ [@Frery1997], the $\Gamma_{\text{SAR}}$  model is a particular case of the $\mathcal{G}^0_I$ distribution. Specifically, for a given $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) 
$$
when $\alpha\to-\infty$.

@@

Introduced by Alfréd Rényi in 1961\ [@renyi1961measures], this measure generalizes several well-known entropies, including the Shannon entropy\ [@Ribeiro2021]. For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda$, with $\lambda > 0$ and $\lambda \neq 1$, is defined as:
\begin{align}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{align}
Using \eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L) \\  + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr]
\end{multline}
and
\begin{multline}\label{eq-HGI0}
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr)=\ln\mu-\ln L +\ln(-1 - \alpha)\\
+ \frac{1}{\,1 - \lambda\,}
\Bigl[
   \lambda\bigl(
      \ln\Gamma(L - \alpha)
      -\ln\Gamma(-\alpha)
      -\ln\Gamma(L)
   \bigr)\\
   +\ln\Gamma\bigl(\lambda(L - 1) + 1\bigr)
   +\ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr)\\
   -\ln\Gamma\bigl(\lambda(L - \alpha)\bigr)
\Bigr].
\end{multline}
@fig-plot presents the Rényi entropy of $\mathcal{G}^0_I$ as a function of $\mu$ for different $\alpha$ values. As $\alpha \to -\infty$, it approaches the Rényi entropy of $\Gamma_{\text{SAR}}$, which is aligned with the fact that $\Gamma_{\text{SAR}}$ is a limiting case of the $\mathcal{G}^0_I$ model.

```{r fig-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="45%",  fig.pos="hbt", fig.cap="$H_{\\lambda}(\\mathcal{G}^0_I)$ converges to the $H_{\\lambda}(\\Gamma_{\\text{SAR}})$ when $\\alpha\\to-\\infty$, with $L=8$ and $\\lambda=0.8$.", fig.width=5, fig.height=4.0}

@@

For $\lambda \to 1$, this expression converges to the Shannon entropy. As we will see later, this property can lead to improved performance for heterogeneity detection compared to Shannon entropy.

## Estimation of Rényi Entropy

Nonparametric estimation of $H(Z)$ has been widely studied using spacing-based estimators, which rely on differences between order statistics [@vasicek1976test; @Ebrahimi1994;@Wieczorkowski1999; @AlOmari2019]. Recently, Al-Labadi et al.\ [@AlLabadi2024] proposed a nonparametric estimator for Rényi entropy following this approach.

<!-- Let $\bm{Z} = (Z_1, Z_2, \dots, Z_n)$ be a random sample from $F(z)$ with order statistics $Z_{(1)} \leq Z_{(2)} \leq \dots \leq Z_{(n)}$. The $m$-spacing density estimator is defined as: -->
Let $\bm{Z}=(Z_1, Z_2,\ldots,Z_n)$ be a random sample from the distribution $F(z)$ and denote its order statistics by $Z_{(1)} \leq Z_{(2)} \leq \dots \leq Z_{(n)}$.
The density estimator based on $m$-spacing is defined as:
$$
f_n(Z_{(i)}) = \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}},
$$
where $Z_{(i-m)} = Z_{(1)}$ when $i \leq m$, and $Z_{(i+m)} = Z_{(n)}$ if $i \geq n - m$. 
The coefficient $c_i$ is given by:
$$
c_i = 
\begin{cases}
\frac{m + i - 1}{m}, & \text{if } 1 \leq i \leq m, \\%[6pt]
2, & \text{if } m+1 \leq i \leq n - m, \\%[6pt]
\frac{n + m - i}{m}, & \text{if } n - m + 1 \leq i \leq n.
\end{cases}
$$

@@

## Optimal $\lambda$ for small sample size

We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator, specifically for a sample size of $n=49$.
To identify this optimal value, we analyze both the mean squared error (MSE) and bias of the estimator across different values of $\lambda$. Lower MSE and bias indicate better performance of the estimator in approximating the true entropy.

Based on the results, we find that the optimal value is $\lambda = 0.9$, as it minimizes the MSE while maintaining a low bias. This choice is consistent with other values of $L > 1$. However, for $L = 1$, the optimal $\lambda$ tends to be higher (e.g., $\lambda = 3$) to achieve good results. @fig-plotf illustrates the case for $L = 5$.

```{r fig-plotf, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="45%",  fig.pos="hbt", fig.cap="Bias and MSE as a function of $\\lambda$, with $n=49$, $L=5$.", fig.width=8, fig.height=6}

@@

Following Refs.\ [@Frery2024;@Alpala2024], we also refine the non-parametric entropy estimator $\widehat{H}_{\lambda}$ using the bootstrap technique and apply the improved version:
$$
\widetilde{H}_{\lambda} = 2\widehat{\theta}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{\theta}_b(\bm{Z}^{(b)}),
$$
where $B$ is the number of new samples generated through resampling with replacement.

A Monte Carlo study with $1000$ samples from $\Gamma_{\text{SAR}}$ ($\mu = 1$, $L=5$, $\lambda = 0.9$) confirms that this approach, using $B = 200$ replications, reduces bias and MSE for small sample sizes, as shown in @fig-Plot_bias_msef3.

@@

```{r fig-Plot_bias_msef3, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="45%", fig.pos="hbt",  fig.cap="Bias and MSE of the Rényi entropy estimator for the $\\Gamma_{\\text{SAR}}$, with $\\lambda=0.9$, $\\mu=1$ and $L=5$.", fig.width=8, fig.height=7}

@@

## Hypothesis Test Rationale

Our approach relies on constructing a test statistic that compares a non-parametric estimate of the Rényi entropy to the theoretical value of the Rényi entropy under the null hypothesis (i.e., a fully developed speckle scenario modeled by $\Gamma_{\text{SAR}}$). 
We aim to test the following hypotheses: 
$$
\begin{cases}\mathcal{H}_0: \widehat{H}^*= H_{\lambda}(\Gamma_{\text{SAR}})\\ 
  \mathcal{H}_1:\widehat{H}^*= H_{\lambda}(\mathcal{G}^0_I).\end{cases}
$$

If the difference between the estimated and theoretical entropy values is statistically significant, we reject the null hypothesis and conclude that the data do not follow the Gamma model, indicating the presence of heterogeneity.

@@

Assume we have an estimator $\widehat{H}_{\lambda}$ for the Rényi entropy of an arbitrary model. In the case of the $\Gamma_{\text{SAR}}$ model, this estimator is expected to be close to the theoretical expression given in Equation \eqref{eq-HGammaSAR}. 
Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S(\bm{Z}; L) = \widehat{H} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\  
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\}.
\end{multline}
When the null hypothesis holds, this test statistic is expected to be close to zero.

For illustrative purposes, @fig-density_entropyR provide the empirical density of $S(\bm{Z}; L)$  for the $\Gamma_{\text{SAR}}$ model, with $\lambda=0.9$ and $L \in \{5,18\}$.
For subsequent simulations we use the $\widetilde{H}_{\lambda}$ estimator.

```{r Simulated_densityR, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

```{r fig-density_entropyR, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="47%", fig.pos="hbt",  fig.cap="Empirical densities obtained from $S_{\\widetilde{H}_{\\lambda}}(\\bm{Z}; L)$ test under the null hypothesis.", fig.width=9, fig.height=5.0}

@@

Under $\mathcal{H}_0$, the distribution of the test statistic $S$ is asymptotically normal. Thus, due to the asymptotic properties of the entropy estimator, and assuming the sample is sufficiently large, the statistic follows a standard normal distribution.

Therefore, the $p$-value is computed as $2\Phi(-\mid \varepsilon \mid)$,
where $\Phi$ is the standard Gaussian cumulative distribution function, and $\varepsilon$ is given by:  
$$
\varepsilon = \frac{\widetilde{H}_i - \widehat{H}_i(\Gamma_{\text{SAR}})}{\widehat{\sigma}_{i}}.
$$
An important aspect of hypothesis testing is ensuring that the test maintains the expected type I error rate (size) while achieving high power (sensitivity to departures from $\mathcal{H}_0$).
To assess these properties, we performed a Monte Carlo simulation with $1000$ replications at significance levels  $1\%$, $5\%$, and $10\%$, evaluating the test under the null hypothesis ($\Gamma_{\text{SAR}}$ distribution) and varying sample sizes, values of $L$, and $\mu=1$. 
The observed type I error rates aligned with the nominal values, confirming the test validity.

Additionally, we examined the power under the alternative hypothesis ($\mathcal{G}^0_I$ distribution) with $\mu=1$, $\lambda=0.9$ and $\alpha=-2$.
As expected, power improves as the sample size and number of looks increase the effectiveness demonstrating of the test. The results are shown in Table \ref{tab:table_size_power}.

>>>>

---
title: Rényi Entropy-Based Heterogeneity Detection in SAR Data
format:
  ieee-pdf:
    pdf-engine: pdflatex # xelatex
    keep-tex: true  
    #conference: true # comment this line to use journal
    #journaltype: conference # comment this line to use journal
    fig-cap-location: bottom # to crossref figure
    link-citations: true
    colorlinks: true
    linkcolor: black # equations
    citecolor: black # cites
    urlcolor: black
  #ieee-html: default

author:
  #- id: 
  - name: Janeth Alpala, Abraão D.\ C.\ Nascimento 
    affiliations:
      - name: Universidade Federal de Pernambuco
        department: Departamento de Estatística
        city: Recife 
        country: PE, Brazil
        postal-code: 50670-901
      #- name: Unknown affiliation
   # orcid: 
    email: janeth.alpala@ufpe.br, abraao@de.ufpe.br
    #url: 
    #membership: "Member, IEEE"
    #attributes:
    #  corresponding: true
    #photo: 
    #bio: |
    #  Use `IEEEbiography`  with figure as  option and
    #  the author name as the argument followed by the biography text.
  - name: Alejandro C.\ Frery
    affiliations:
      - name: Victoria University of Wellington
        department: School of Mathematics and Statistics
        city: Wellington 
        country: New Zealand
        postal-code: 6140
      #- name: Unknown affiliation
    #orcid: 
    email: alejandro.frery@vuw.ac.nz
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems. A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes. In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed. 
  In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task. 
  In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy. 
  In particular, we use a nonparametric estimator for the Rényi entropy and investigate some of its properties. 
  As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories. 
  The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy.
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR,  hypothesis tests]

#funding: 
 # statement: "The `quarto-ieee` "
pageheader:
  left: Journal XXX, Month Year
  right: #'D. Folio:  A Sample Article Using quarto-ieee'

header-includes:
   - \usepackage[english]{babel}
   - \usepackage{bm,bbm}
   - \usepackage{mathrsfs}
   - \usepackage{siunitx}
   - \usepackage{graphicx}
   - \usepackage{url}
   - \usepackage[T1]{fontenc}
   #- \usepackage{polski}
   - \usepackage{booktabs}
   - \usepackage{color}
   - \usepackage{mathtools}
  # - \usepackage[utf8]{inputenc}
   - \usepackage{hyperref}
   #- \hypersetup{draft} #Desactiva enlaces y referencias cruzadas
   - \usepackage{float}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{xcolor}
   - \usepackage{amsmath}
   - \usepackage{hyperref}
   - \hypersetup{
      colorlinks=true,
      linkcolor=black,
      filecolor=black,
      citecolor=black,
      urlcolor=black}

#bibliography: references.bib
bibliography: ../../Common/references.bib

@@

---
```{r setup, include=FALSE}

@@

# Introduction
[S]{.IEEEPARstart}[ynthetic]{}
 Aperture Radar (SAR) technology has become essential for a wide range of applications\ [@Yu2023;@Akbarizadeh2012;@Mondini2021]. It provides high-resolution data that is independent of sunlight and operates in a variety of weather conditions to facilitate global Earth monitoring\ [@Zeng2020].
Despite its advantages, the effective use of SAR data depends on understanding its statistical properties because this data is affected by speckle, a noise-like interference effect\ [@Argenti2013; @Choi2019]. In intensity format, speckle is typically non-Gaussian, and its presence complicates subsequent image analysis tasks.

The $\mathcal{G}^0_I$ distribution has been shown to be an effective law for SAR intensity data, as it captures different levels of roughness. A special case is the Gamma distribution, which arises when speckle is fully developed, indicating homogeneous regions. 
Although these assumptions offer flexibility in describing SAR intensities, selecting the appropriate distribution can be challenging for two main reasons: the small sample sizes used in practical applications and the inherent difficulties associated with parameter estimation. 
These challenges complicate model selection and highlight the need to explore alternative statistical approaches.

Entropy measures have gained attention as valuable statistical tools for analyzing SAR data, with applications in edge detection\ [@Nascimento2014], segmentation\ [@Nobre2016], classification\ [@Cassetti2022], and noise reduction\ [@Chan2022].
Traditionally, Shannon entropy\ [@Shannon1948] has been widely used to quantify uncertainty and disorder in data. However, this study explores an alternative information measure: Rényi entropy—a generalization of Shannon’s formulation. 
This more flexible approach provides additional insights for identifying heterogeneity, making it a promising tool for enhancing SAR image analysis.

@@

The main distributions considered for SAR intensity data are the $\Gamma_{\text{SAR}}$ distribution, which is suitable for fully developed speckle, and the $\mathcal{G}^0_I$ distribution, which is able to describe roughness\ [@Frery1997]. These distributions are characterized by the following probability density functions (pdfs):
\begin{align}
	f_{\Gamma_{\text{SAR}}}\bigl(z;L, \mu \bigr) 
    &= \frac{L^L}{\Gamma(L)\,\mu^L} z^{L-1} 
    \exp \biggl(-\frac{Lz}{\mu}\biggr)
    \mathbbm 1_{\mathbbm R_+}(z) \label{E:gamma1}
    \intertext{and}
    f_{\mathcal{G}^0_I}\bigl(z; \mu, \alpha, L \bigr) 
    &= \frac{L^L\,\Gamma(L-\alpha)}
    {\bigl[-\mu(\alpha+1)\bigr]^{\alpha} \Gamma(-\alpha)\,\Gamma(L)}
    \notag \\
    &\quad \times
    \frac{z^{L-1}}
    {\bigl[-\mu(\alpha+1)+Lz\bigr]^{L-\alpha}}
    \mathbbm 1_{\mathbbm R_+}(z), \label{E:gi01}
\end{align}
where $\mu > 0$ is the mean,
$\alpha < 0$ measures the roughness, $L \geq 1$ is the number of
looks, $\Gamma(\cdot)$ is the gamma function, and
$\mathbbm 1_{A}(z)$ is the indicator function of the set $A$. 
As demonstrated by\ [@Frery1997], the $\Gamma_{\text{SAR}}$  model is a particular case of the $\mathcal{G}^0_I$ distribution. Specifically, for a given $\mu$ fixed,
$$
f_{\mathcal{G}^0_I}\big(z; \mu, \alpha, L\big)
\longrightarrow 
f_{\Gamma_{\text{SAR}}}(z;L, \mu) 
$$
when $\alpha\to-\infty$.

@@

Introduced by Alfréd Rényi in 1961\ [@renyi1961measures], this measure generalizes several well-known entropies, including the Shannon entropy\ [@Ribeiro2021]. For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda$, with $\lambda > 0$ and $\lambda \neq 1$, is defined as:
\begin{equation}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{equation}
Using \eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L) \\  + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr]
\end{multline}
and
\begin{multline}\label{eq-HGI0}
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr)=\ln\mu-\ln L +\ln(-1 - \alpha)\\
+ \frac{1}{\,1 - \lambda\,}
\Bigl[
   \lambda\bigl(
      \ln\Gamma(L - \alpha)
      -\ln\Gamma(-\alpha)
      -\ln\Gamma(L)
   \bigr)\\
   +\ln\Gamma\bigl(\lambda(L - 1) + 1\bigr)
   +\ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr)\\
   -\ln\Gamma\bigl(\lambda(L - \alpha)\bigr)
\Bigr].
\end{multline}
The derivation of this result is given in the Appendix.

@fig-plot presents the Rényi entropy of $\mathcal{G}^0_I$ as a function of $\mu$ for different $\alpha$ values. As $\alpha \to -\infty$, it approaches the Rényi entropy of $\Gamma_{\text{SAR}}$, which is aligned with the fact that $\Gamma_{\text{SAR}}$ is a limiting case of the $\mathcal{G}^0_I$ model.

```{r fig-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="45%",  fig.pos="hbt", fig.cap="$H_{\\lambda}(\\mathcal{G}^0_I)$ converges to the $H_{\\lambda}(\\Gamma_{\\text{SAR}})$ when $\\alpha\\to-\\infty$, with $L=8$ and $\\lambda=0.8$.", fig.width=4.5, fig.height=3.5}

@@

## Nonparametric Estimation of Rényi Entropy

Nonparametric estimation of $H(Z)$ has been widely studied using spacing-based estimators, which rely on differences between order statistics [@vasicek1976test; @Ebrahimi1994; @Wieczorkowski1999; @AlOmari2019]. Recently, Al-Labadi et al.\ [@AlLabadi2024] proposed a nonparametric estimator for Rényi entropy following this approach.

Let $\bm{Z}=(Z_1, Z_2,\ldots,Z_n)$ be an independent and identically distributed random sample of size $n$ from a distribution $F$, and let $Z_{(1)} \leq Z_{(2)} \leq \dots \leq Z_{(n)}$ denote its order statistics.
The $m$-spacing density estimator is defined as:
$$
f_n(Z_{(i)}) = \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}},
$$
where $Z_{(i-m)} = Z_{(1)}$ when $i \leq m$, and $Z_{(i+m)} = Z_{(n)}$ if $i \geq n - m$. 
The coefficient $c_i$ is given by:
$$
c_i = 
\begin{cases}
\frac{m + i - 1}{m}, & \text{if } 1 \leq i \leq m, \\%[6pt]
2, & \text{if } m+1 \leq i \leq n - m, \\%[6pt]
\frac{n + m - i}{m}, & \text{if } n - m + 1 \leq i \leq n.
\end{cases}
$$

@@

## On $\lambda$ optimality for a sample size

We aim to determine the optimal order $\lambda$ for the Rényi entropy estimator for a sample size $n=49$.
To identify this optimal value, we analyze both the mean squared error (MSE) and bias of the estimator across different values of $\lambda$. Lower MSE and bias indicate better performance of the estimator in approximating the true entropy.

Based on the results, we find that the optimal value is $\lambda = 0.9$, as it minimizes the MSE while maintaining a low bias. This choice is the same for $L > 1$. However, for $L = 1$, the optimal $\lambda$ tends to be higher (e.g., $\lambda = 3$) to achieve good results. @fig-plotf illustrates the case for $L = 5$.
```{r fig-plotf, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="42%",  fig.pos="hbt", fig.cap="Bias and MSE as a function of $\\lambda$, with $n=49$, $L=5$.", fig.width=7, fig.height=4.5}

@@

Following Refs.\ [@Frery2024;@Alpala2024], we also refine the non-parametric entropy estimator $\widehat{H}_{\lambda}$ in \eqref{eq:est_R} using the bootstrap technique to reduce its bias and apply the improved version, denoted as $\widetilde{H}_{\lambda}$:
$$
\widetilde{H}_{\lambda} = 2\widehat{H}_{\lambda}(\bm{Z}) - \frac{1}{B} \sum_{b=1}^{B} \widehat{H}_{\lambda}(\bm{Z}^{(b)}),
$$
where $B$ is the number of bootstrap replications, and $\bm{Z}^{(b)}$ is the $b$-th resampled dataset obtained by drawing $n$ observations with replacement from $\bm{Z}$.

A Monte Carlo study with $1,000$ samples from $\Gamma_{\text{SAR}}$ ($\mu = 1$, $L=5$, $\lambda = 0.9$), using $B = 200$ replications, confirms that the bootstrap-corrected estimator $\widetilde{H}_{\lambda}$  outperforms the original $\widehat{H}_{\lambda}$ by reducing both bias and MSE for small sample sizes, as shown in @fig-Plot_bias_msef3.

@@

```{r fig-Plot_bias_msef3, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="42%", fig.pos="hbt",  fig.cap="Bias and MSE of the Rényi entropy estimators for $\\Gamma_{\\text{SAR}}$.", fig.width=7, fig.height=5}

@@

```
For subsequent simulations we use the $\widetilde{H}_{\lambda}$ estimator.

## Hypothesis Testing

We test whether the observed data come from a homogeneous ($\Gamma_{\text{SAR}}$) or a heterogeneous ($\mathcal{G}^0_I$) region, as follows:
\begin{equation}\label{eq:hypothesis_test}
\begin{cases}
\mathcal{H}_0: \mathbb{E}[\widetilde{H}_{\lambda}] = H_{\lambda}(\Gamma_{\text{SAR}}), & \text{(Homogeneous region)} \\[6pt]
\mathcal{H}_1: \mathbb{E}[\widetilde{H}_{\lambda}] = H_{\lambda}(\mathcal{G}^0_I), & \text{(Heterogeneous region)}
\end{cases}
\end{equation}

Under $\mathcal{H}_0$, the expected value of the entropy estimator should match the theoretical $H_{\lambda}(\Gamma_{\text{SAR}})$. If the estimated entropy significantly deviates from its expected value, we reject $\mathcal{H}_0$, indicating the presence of heterogeneity.

Since homogeneity corresponds to $\alpha \to -\infty$ in the $\mathcal{G}^0_I$ model, classical inference methods are inapplicable. Instead, we propose a test statistic based on the Rényi entropy estimator, comparing it with its theoretical expectation under $\mathcal{H}_0$ to distinguish between homogeneous and heterogeneous areas.

@@

Assume we have an estimator $\widetilde{H}_{\lambda}$ for the Rényi entropy of an arbitrary model. For testing \eqref{eq:hypothesis_test}, this estimator is expected to be close to the theoretical expression given in Equation \eqref{eq-HGammaSAR}. 
Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\  
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\}.
\end{multline}
where $\ln \widehat{\mu}={\textstyle\frac{1}{n}\sum_{i=1}^n Z_{i}}$ is the natural logarithm of the sample mean.

When the null hypothesis holds, this test statistic is expected to be close to zero.
For illustrative purposes, @fig-density_entropyR shows the empirical density of $S(\bm{Z}; L)$, based on $10,000$ simulations of the $\Gamma_{\text{SAR}}$ model, with $\lambda=0.9$ and $L \in \{5,18\}$.
```{r Simulated_densityR, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

@@

```{r fig-density_entropyR, echo=FALSE, message=FALSE, warning=FALSE, fig.fullwidth = TRUE, out.width="47%", fig.pos="hbt",  fig.cap="Empirical densities of $S_{\\widetilde{H}_{\\lambda}}(\\bm{Z}; L)$ under $\\mathcal{H}_0$.", fig.width=9, fig.height=4.0}

@@

```
Under the asymptotic properties of the entropy estimator\ [@vasicek1976test], for sufficiently large samples, $S(\bm{Z}; L)$ follows an asymptotic normal distribution:
\begin{equation*}
S(\bm{Z}; L) \overset{d}{\longrightarrow} \mathcal{N}(\mu_S,\,\sigma^{2})\,.
\end{equation*}
We define the standardized test statistic:
\begin{equation*}
\varepsilon = \frac{S(\bm{Z}; L) - \hat{\mu}_S}{\hat{\sigma}_S},
\end{equation*}
where $\hat{\mu}_S = \mathbb{E}[S(\bm{Z}; L)]$ and $\hat{\sigma}_S = \sqrt{\text{Var}[S(\bm{Z}; L)]}$.
By the central limit theorem, under the null hypothesis, $\varepsilon$ asymptotically follows a standard normal distribution:
\begin{equation*}
\varepsilon \overset{d}{\longrightarrow} \mathcal{N}(0,1)\,,
\end{equation*}
where $d$ represents convergence in distribution.
@fig-density_entropyR presents the behavior of the empirical distribution of the standardized test statistic.
Thus, the $p$-values are calculated as $2\Phi(-|\varepsilon|)$,
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal distribution.
```{r Simulated_densityR_standardized0, echo=FALSE, message=FALSE}
set.seed(1234567890, kind = "Mersenne-Twister")

R <- 20000
mu <- 1
B <- 100
